### 2013-06-04 paper_draft.tex
added thanks to reviewers
whops, linebreak in abstract
Many minor changes to formulations. Replaced disease by illness. Normalized some spellings. Added short explanatory sentences.

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Genome-Wide Association Studies are widely used to correlate phenotypic traits with genetic variants. These studies usually compare the genetic variation between two groups to single out certain Single Nucleotide Polymorphisms (SNPs) that are linked to a phenotypic variation in one of the groups. However, it is necessary to have a large enough sample size to find statistically significant correlations. Direct-To-Consumer (DTC) genetic testing can supply additional data: DTC-companies offer the analysis of a large amount of SNPs for an individual at low cost without the need to consult a physician or geneticist. Over 100,000 people have already been genotyped through Direct-To-Consumer genetic testing companies. However, this data is not public for a variety of reasons and thus cannot be used in research. It seems reasonable to create a central open data repository for such data, but it was previously unknown if and how people would submit their data to such a repository. Here 
%we present a survey which evaluates whether people are willing to publicly share their genetic information. In the light of those results 
we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish their genetic data at no cost along with phenotypic information. Through this crowdsourced effort of collecting genetic and phenotypic information openSNP has become a 
%valuable 
resource for a wide area of studies, including Genome-Wide Association S
tudies. openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr

@@

The availability of new DNA sequencing techniques has shifted the focus of biological data acquisition towards new biomedical applications.
Many diseases - for example Alzheimer's \cite{alzheimer}, Parkinson's \cite{parkinsons} or different types of cancers \cite{breastcancer,prostatecancer} - are at least partially heritable, so the genome of patients can be used for diagnostic purposes. Using the genetic information of patients for diagnostics is made possible through the sharp decrease in costs for analysing genetic information \cite{Brown1999}.

@@

If genetic information on more than one individual is known, the analysis of
allele frequencies of Single Nucleotide Polymorphisms (SNPs) can be used to associate such SNPs with diseases and other inheritable traits. Genome-Wide Association Studies
(GWAS) make use of statistics to compare the allele frequencies in patients to the alleles in healthy controls. This
enables GWAS to find SNPs which are significantly overrepresented in patients and associates those SNPs with a trait or disease.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising. To date, over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different diseases and traits \cite{Hindorff2009}.

GWAS are not only performed inside the traditional scientific community. 
Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
receive an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
alone had over 100,000 customers \cite{23andMe2011}
- the company realizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
traits and diseases. With the consent of the customer, the data is used for association studies. 23andMe has published several 
articles in which known findings are replicated together with new associations disorders like Parkinson's Disease \cite{Eriksson2010, Do2011}. 
So far, over 30,000 23andMe-customers have participated in 23andMe's association studies, which proves that this data source has a lot of potential for other researchers.

@@

%Enter here: definition, benefits of sourcing
Crowdsourcing, giving a task into the hands of a potentially large number of - mainly intrinsically motivated - people, has become a widely used practice in the internet age and is getting adopted in the realm of science as well. One of the main benefits of crowdsourcing is that small contributions to a project pile up to create a larger work, which would have been virtually impossible to create otherwise. This especially benefits scientists who might not have enough funding or time to create data. Or in cases where the amounts of data are too large to be analyzed by researchers alone. % added this, because neither galaxy zoo or foldit crowdsource data creating but only the analysis
 Galaxy Zoo and FoldIt \cite{Eiben2012, GalaxyZoo} are two of the best known examples. 
Galaxy Zoo enabled thousands of amateur astronomers to walk through thousands of telescopic images to categorize the shown objects at a rate which could not have been matched by the efforts of professional astronomers.
Similarly crowdsourcing can not only be applied to analyzing data, but also to collecting data. This approach has been shown to work when it comes to tracking bird migration \cite{CrowdsourcingReview2010}. With the advent of DTC genetic testing and the internet a similar approach can now be brought to human genetics.

There have been studies investigating how likely customers of such companies are to share their data. \cite{Darst2013} investigated the likelihood of 2,024 individuals to share their test results with their health-care providers and found that 26.5\% (540 individuals) did share their results with physician or health-care provider. Those that shared where older, had a higher income and were less concerned about testing or the privacy implications of this sharing compared to customers that didn't share their data.

@@

\subsection*{Sharing genotypic information}
Currently users can upload their genotyping results from the companies \textit{23andMe}, \textit{deCODEme} and \textit{FamilyTreeDNA} via a web interface to the openSNP 
project. There is experimental support for uploading exomes in the VCF format \cite{Danecek01082011}, as \textit{23andMe} recently started exome sequencing for its customers. So far only the SNPs of the exome data sets are stored on openSNP, but the downloads include all variation found in the exome.
The uploaded data is published under the Creative Commons Zero license, 
which - in accordance with the Panton Principles \cite{10.1371/journal.pbio.1001195} - 
allows a complete re-use of the data without any constraints.
Between the launch of openSNP on 09/27/2011 and 10/27/2012, 633 people have signed 
up with openSNP, and 270 genetic datasets have been made available. As of 10/27/2012, the openSNP database lists 215,546,685 genotypes which are distributed over 2,140,643 unique SNPs.
Figures \ref{Figure1_label} and \ref{Figure2_label} depict the increase in users and genotyping files since September 2011.

@@

\subsection*{Connection to external services}
In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. 
A total number of 21,134 documents relevant to the SNPs listed in openSNP could be found in the publication and annotation databases of Mendeley, the Public Library of Science, in the \emph{GET Evidence System} \cite{Ball24072012} and the \emph{NHGRI GWAS Catalog} \cite{Hindorff2009} and in the crowdsourced SNPedia (Figure \ref{Figure3_label}).
Of the primary literature listed on Mendeley, the \emph{NHGRI GWAS Catalog} \& the Public Library of Science, about 20 \% are released in open access journals and can be accessed free of charge (Figure \ref{oa_label}), although probably not all publications on Mendeley are correctly flagged and the \emph{NHGRI GWAS Catalog} does not give details on whether a publication is Open Access or not. So the total number of Open Access publications might be higher.

For usability reasons, 
SNPs are ranked by the amount of information gathered through the external services. The external services themselves are ranked by how easily non-scientists can understand information 
from these sources and how available this information is to the public. The SNPedia entries are given the highest impact, as those are already manually curated and summarized in plain English, followed by open access publications out of 
the Public Library of Science and the curated databases of the \emph{GET Evidence System} and the \emph{NHGRI GWAS Catalog}. Lowest values are given to the Mendeley results, as the publications listed there are for the most part not freely available without subscriptions or one-time payments. 
An entry on SNPedia is valued 2.5 times as high as a PLoS publication or entries in \emph{GET} or the \emph{GWAS Catalog} and 5 times as high as a Mendeley entry.

@@

\subsection*{Data access}
openSNP offers complete access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypic variation in a comma-separated table. The genetic data is also 
accessible through the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}, which offers all data for specific chromosomes and specific positions on single chromosomes. 
An example of how the DAS can be used is implemented on openSNP, where users' genotypes are visualized inside a genome browser. All chromosomal positions are based on the human reference genome NCBI37, as this is the standard reference used by DTC providers right now.

@@

\subsection*{Other crowdsourcing platforms}
Projects similar to openSNP are the SNPedia, the Personal Genome Project and PatientsLikeMe.com (see table \ref{tab:platforms} for an overview). The focus of the SNPedia is the aggregation and summary of primary scientific literature on SNPs. The project uses a Wiki to store and display the data collected by volunteers contributing to the project. The data is mainly organized by the unique Rs-ID, as given by dbSNP. If Rs-IDs are missing the identifiers given by the DTC testing companies may be used, similar to the way openSNP stores the data. For individual SNPs pages may list scientific literature and summaries on the found impact can be given. As those pages are largely created manually and not automatically through database access these summaries may not be complete. openSNP utilizes the SNPedia in that it crawls their data for SNPs, the summary of the impact and the magnitude a SNP has. While they offer a page listing download-URLs the SNPedia does not offer any uploading capabilities for genetic data and has no APIs to easily access SNPs or data subsets in the different data sets. Similarly there is no way for users of SNPedia to share their phenotypes in a machine readable format.

@@

PatientsLikeMe is a community for patients with life-changing illnesses to track and share the development of their illness with other patients with similar illnesses \cite{Wicks2010}.  This helps patients in understanding their diseases better - 72\% of surveyed participants found the site "moderately" or "very helpful", for example when it comes to starting a new medication (37\% found the site helpful), or when it comes to changing the medication (27\%). Some subsets of data stored in PatientsLikeMe are open to the public and has been shown to be useful for research, for example in Multiple Sclerosis \cite{Bove2013}. Alternatively access to the data they store can be licensed by researchers for a fee.

@@

While we try to suggest similar entries to the users, 
there are some cases where users will not follow those suggestions, so duplicates or similar phenotypes or variations in traits may arise. There are two possible solutions to this problem: The first one would be to only allow a trusted subset of users to enter new phenotypes. The other one 
would be to make users enter all possible variations of a phenotype while creating a new phenotype, so that later users cannot add 
variations that have not been available from the start.

In both cases it makes it harder 
for users to enter their data which raises the bar for participation.
We decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS with the data to perform additional quality control.

@@

Furthermore, it is impossible to verify whether users who have uploaded data are actually the sources for that data. This opens the venue to potentially malicious usage, as genotypings from strangers can be uploaded, as well as misinformation about phenotypes can be entered. The openSNP project has currently no means to verify the validity of data given to us by users. Of course, users can always delete their data or contact the team to delete stored data. No copies are kept at any time.

@@

\subsection*{Technical implementation of the platform}
The main platform is implemented using the web framework Ruby on Rails 3.0.10. Postgres 9.2 is used as the main database backend for Rails. 
The database stores genotyping results, users' phenotypic information, literature results from Mendeley and the Public Library of Science as well as summaries on SNPs 
which can be found in SNPedia. The literature database of Mendeley is queried using the REST API, which delivers results in JSON. The literature database of 
the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format. Summaries on SNPs are provided by SNPedia, 
through querying the content via the MediaWiki API. The \emph{NHGRI GWAS Catalog} and the \emph{GET Evidence System} provide complete dumps in plain text formats. Those are regularly downloaded and parsed. SNPs that are described as 'Insufficiently evaluated' in the \emph{GET Evidence System} are not stored. All databases are queried or parsed using the unique identifier of each SNP as the search term.

@@

A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in Figure \ref{Figure4_label}. The source code of openSNP is 
published under the MIT license and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
We thank Dr. Manuel Corpas and Prof. David Edwards for constructive advice in grammar, spelling and structure of this study. Further thanks go to Samantha Clark, Fabian Zimmer and Dan Bolser for providing valuable feedback and feature ideas during the development and Thomas Down, author of \emph{BioDalliance}, and Rafael Jimenez, author of \emph{MyKaryoView}, for their support on implementing the Distributed Annotation System and the genome browser. For help with their APIs we are grateful to Mike Cariaso of \emph{SNPedia} and the PLOS \& Mendeley API teams. We would especially like to thank the users of openSNP.org for their participation, their constructive criticism and bug-finding abilities and especially for sharing their genotyping and phenotype data.

@@

\begin{figure}[!ht]
	\begin{center}
		\includegraphics[scale=0.6]{25_10_2012_Graphs/new_pie.png}
	\end{center}
	\caption{
	{\bf Ratio of Open Access Publications.} Green pieces are Open Access. The \emph{NHGRI GWAS Catalog} doesn't give information about the Open Access status.}
	\label{oa_label}
\end{figure}

@@

\section*{Tables}
\begin{table}
\caption{Comparison of crowdsourced genetics platforms.}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
Name & Provides  & Provides & Open  & API & IRB approval & License \\ 
& annotation & data & enrollment & & & \\
\hline
SNPedia & \checkmark & \xmark & N/A & \checkmark & N/A & CC-NC-SA 3.0 \\ 
\hline
PGP & \checkmark & \checkmark & \xmark & \xmark & \checkmark & CC-BY 1.0, CC0 \\
\hline
PatientsLikeMe & \xmark & \checkmark & \checkmark & \xmark & \checkmark & Closed, CC-BY-SA 3.0 \\
\hline
openSNP & \checkmark & \checkmark & \checkmark & \checkmark & \xmark & CC-BY 3.0 \\
\hline
\end{tabular}
\label{tab:platforms}
\end{table}
%\begin{table}
%\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped. }  
%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%\hline
%& Mean \emph{Genotyped} & Mean \emph{Not Genotyped} & Mean Difference & Standard Error\\ 
%\hline
%\textbf{Motivation for sharing data in participants who are already genotyped} & & & & \\
%\hline 
%... curious & 3.82 & 2.66 & 1.159 & 0.193 \\ \hline % checked
%... want to help scientists & 4.64 & 4.18  & 0.465 & 0.128 \\ \hline % checked
%... for personal benefits & 3.77 & 3.32  & 0.448 & 0.183 \\ \hline % checked
%\textbf{Motivation for not sharing in participants who are not planning to get genotyped} & & & & \\ \hline
%... fear of discrimination & 3.09 & 4.15  & 1.06 & 0.195 \\ \hline % checked
%... breach of privacy & 3.01 & 3.68  & 0.666 & 0.211 \\ \hline % checked
%... fear of personalized advertising & 3.03 & 3.88  & 0.848 & 0.208 \\ \hline % checked
%... negative consequences for family members & 2.93 & 3.57  & 0.639 & 0.197 \\ \hline 
%\end{tabular}
%\label{tab:motivations1}
%\end{table}
%
%\begin{table}
%\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with participants who would not share their data.}
%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%\hline
%& Mean {Sharing} & Mean {Not Sharing} & Mean Difference & Standard Error \\ 
%\hline
%\textbf{Motivation for sharing genotypings in participants who would share} & & & & \\
%\hline
%... curiosity & 3.86 & 1.87 & 1.99 & 0.321 \\ \hline % checked
%... want to help science & 4.70 & 3.13 & 1.57 & 0.199 \\ \hline % checked
%... for personal benefits & 3.68 & 2.73 & 0.951 & 0.308 \\ \hline % checked
%\textbf{Motivation for sharing genotypings in participants who would not share} & & & & \\
%\hline
%... fear of discrimination & 3.21 & 4.73 & 1.52 & 0.322 \\ \hline % checked
%... fear of consequences for family members & 2.79 & 3.93 & 1.146 & 0.32 \\ \hline % checked
%... fear of personalized advertising & 3.17 & 4.29 & 1.112 & 0.357 \\ \hline % checked
%\end{tabular}
%\label{tab:motivations2}
%\end{table}
%\begin{table}[!ht]
%\caption{
%\bf{Table title}}
%\begin{tabular}{|c|c|c|}
%table information
%\end{tabular}
%\begin{flushleft}Table caption
%\end{flushleft}
%\label{tab:label}
% \end{table}

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Genome-Wide Association Studies are widely used to correlate phenotypic traits with genetic variants. These studies usually compare the genetic variation between two groups to single out certain Single Nucleotide Polymorphisms (SNPs) that are linked to a phenotypic variation in one of the groups. However, it is necessary to have a large enough sample size to find statistically significant correlations. Direct-To-Consumer (DTC) genetic testing can supply additional data: DTC-companies offer the analysis of a large amount of SNPs for an individual at low cost without the need to consult a physician or geneticist. Over 100,000 people have already been genotyped through Direct-To-Consumer genetic testing companies. However, this data is not public for a variety of reasons and thus cannot be used in research. It seems reasonable to create a central open data repository for such data.
%, but it was previously unknown if and how people would submit their data to such a repository.
%we present a survey which evaluates whether people are willing to publicly share their genetic information. In the light of those results 
Here we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish their genetic data at no cost along with phenotypic information. Through this crowdsourced effort of collecting genetic and phenotypic information openSNP has become a 
%valuable 
resource for a wide area of studies, including Genome-Wide Association Studies. openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr

@@

The availability of new DNA sequencing techniques has shifted the focus of biological data acquisition towards new biomedical applications.
Many illnesses - for example Alzheimer's \cite{alzheimer}, Parkinson's \cite{parkinsons} or different types of cancers \cite{breastcancer,prostatecancer} - are at least partially heritable, so the genome of patients can be used for diagnostic purposes. Using the genetic information of patients for diagnostics is made possible through the sharp decrease in costs for analysing genetic information \cite{Brown1999}.

@@

If genetic information on more than one individual is known, the analysis of
allele frequencies of Single Nucleotide Polymorphisms (SNPs) can be used to associate such SNPs with illnesses and other inheritable traits. Genome-Wide Association Studies
(GWAS) make use of statistics to compare the allele frequencies in patients to the alleles in healthy controls. This
enables GWAS to find SNPs which are significantly overrepresented in patients and associates those SNPs with a trait or illness.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising. To date, over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different illnesses and traits \cite{Hindorff2009}.

GWAS are not only performed inside the traditional scientific community. 
Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
receive an analysis of the results, as well as a raw file that includes the customer's individual genotypes. In 2011, 23andMe 
alone had over 100,000 customers \cite{23andMe2011}
- the company realizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
traits and illnesses. With the consent of the customer, the data is used for association studies. 23andMe has published several 
studies in which known findings are replicated together with new associations for disorders like Parkinson's Disease \cite{Eriksson2010, Do2011}. 
So far, over 30,000 23andMe-customers have participated in 23andMe's association studies, which proves that this data source has a lot of potential for other researchers.

@@

%Enter here: definition, benefits of sourcing
Crowdsourcing, giving a task into the hands of a potentially large number of - mainly intrinsically motivated - people, has become a widely used practice in the internet age and is getting adopted in the realm of science as well. One of the main benefits of crowdsourcing is that small contributions to a project pile up to create a larger work, which would have been virtually impossible to create otherwise. This approach especially benefits scientists who might not have enough funding or time to create data, or in cases where the amounts of data are too large to be analyzed by researchers alone.
 Galaxy Zoo and FoldIt \cite{Eiben2012, GalaxyZoo} are two of the best known examples. 
Galaxy Zoo enables amateur astronomers to walk through telescopic images to categorize the shown objects at a rate which could not have been matched by the efforts of professional astronomers.
Similarly crowdsourcing can not only be applied to analyzing data, but also to collecting data. This approach has been shown to work when it comes to tracking bird migration \cite{CrowdsourcingReview2010}. With the advent of DTC genetic testing and the internet a similar approach can now be applied to human genetics.

There have been studies investigating how likely customers of such companies are to share their data. \cite{Darst2013} investigated the likelihood of 2,024 individuals to share their test results with their health-care providers and found that 26.5\% (540 individuals) did share their results with their physician or health-care provider. Those that shared where older, had a higher income and were less concerned about testing or the privacy implications of this sharing compared to customers that didn't share their data.

@@

\subsection*{Sharing genotypic information}
Currently users can upload their genotyping results from the companies \textit{23andMe}, \textit{deCODEme} and \textit{FamilyTreeDNA} via a web interface to the openSNP 
project. There is experimental support for uploading exomes in the VCF format \cite{Danecek01082011}, as \textit{23andMe} recently started exome sequencing for its customers. Due to space constraints openSNP currently only stores the SNPs of the exome data sets.
The uploaded data is published under the Creative Commons Zero license, 
which - in accordance with the Panton Principles \cite{10.1371/journal.pbio.1001195} - 
allows a complete re-use of the data without any constraints.
Between the launch of openSNP on 09/27/2011 and 10/27/2012, 633 people have signed 
up with openSNP, and 270 genetic datasets have been made available. As of 10/27/2012, the openSNP database lists 215,546,685 genotypes which are distributed over 2,140,643 unique SNPs.
Figures \ref{Figure1_label} and \ref{Figure2_label} depict the increase in users and genotyping files since September 2011.

@@

\subsection*{Connection to external services}
In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. 
A total number of 21,134 documents relevant to the SNPs listed in openSNP could be found in the publication and annotation databases of Mendeley, the Public Library of Science, in the \emph{GET Evidence System} \cite{Ball24072012} and the \emph{NHGRI GWAS Catalog} \cite{Hindorff2009} and in the crowdsourced SNPedia (Figure \ref{Figure3_label}).
Of the primary literature listed on Mendeley, the \emph{NHGRI GWAS Catalog} \& the Public Library of Science, about 20 \% are released in open access journals and can be accessed free of charge (Figure \ref{oa_label}), although probably not all publications on Mendeley are correctly flagged and the \emph{NHGRI GWAS Catalog} does not give details on whether a publication is open access or not. So the total number of open access publications might be higher.

For usability reasons, 
SNPs are ranked by the amount of information gathered through the external services. The external services themselves are ranked by how easily non-scientists can understand information 
from these sources and how available this information is to the public. The SNPedia entries are given the highest impact, as those are already manually curated and summarized in plain English, followed by open access publications out of 
the Public Library of Science and the curated databases of the \emph{GET Evidence System} and the \emph{NHGRI GWAS Catalog}. Lowest values are given to the Mendeley results, as the publications listed there are for the most part not freely available without subscriptions or one-time payments. 
An entry on SNPedia is valued 2.5 times as high as a PLOS publication or entries in \emph{GET} or the \emph{GWAS Catalog} and 5 times as high as a Mendeley entry.

@@

\subsection*{Data access}
openSNP offers complete access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypic variation in a comma-separated table. For privacy reasons openSNP does not log any IPs. The genetic data is also 
accessible through the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}, which offers all data for specific chromosomes and specific positions on single chromosomes. 
An example of how the DAS can be used is implemented on openSNP, where users' genotypes are visualized inside a genome browser. All chromosomal positions are based on the human reference genome NCBI37, as this is the standard reference used by DTC providers right now.

@@

\subsection*{Comparing openSNP to other crowdsourcing platforms}
Projects similar to openSNP are the SNPedia, the Personal Genome Project and PatientsLikeMe.com (see table \ref{tab:platforms} for an overview). The focus of the SNPedia is the aggregation and summary of primary scientific literature on SNPs. The project uses a Wiki to store and display the data collected by volunteers contributing to the project. The data is mainly organized by the unique Rs-ID, as given by dbSNP. If Rs-IDs are missing the identifiers given by the DTC testing companies may be used, similar to the way openSNP stores the data. For individual SNPs pages may list scientific literature and summaries on the found impact can be given. As those pages are largely created manually and not automatically through database access these summaries may not be complete. openSNP utilizes the SNPedia in that it crawls their data for SNPs, the summary of the impact and the magnitude a SNP has. While they offer a page listing download-URLs the SNPedia does not offer any uploading capabilities for genetic data and has no APIs to easily access SNPs or data subsets in the different data sets. Similarly there is no way for users of SNPedia to share their phenotypes in a machine readable format.

@@

PatientsLikeMe is a community for patients with life-changing illnesses to track and share the development of their illness with other patients with similar illnesses \cite{Wicks2010}.  This helps patients in understanding their illnesses better - 72\% of surveyed participants found the site "moderately" or "very helpful", for example when it comes to starting a new medication (37\% found the site helpful), or when it comes to changing the medication (27\%). Some subsets of data stored in PatientsLikeMe are open to the public and has been shown to be useful for research, for example in Multiple Sclerosis \cite{Bove2013}. Alternatively access to the data they store can be licensed by researchers for a fee.

@@

While we try to suggest similar entries to the users, 
there are some cases where users will not follow those suggestions, so duplicates or similar phenotypes or variations in traits may arise. There are three possible solutions to this problem: The first one would be to only allow a trusted subset of users to enter new phenotypes. The second one 
would be to make users enter all possible variations of a phenotype while creating a new phenotype, so that later users cannot add 
variations that have not been available from the start. The third one is to exclude users from the phenotype-creation process by allowing users to select their phenotypes from a pre-given set of possible variations.

In the first two cases it makes it harder 
for users to enter their data which raises the bar for participation, and the third case doesn't let users participate at all.
We decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS with the data to perform additional quality control.

@@

Furthermore, it is impossible to verify whether users who have uploaded data are actually the sources for that data. This opens the venue to potentially malicious usage, as genotypings from strangers can be uploaded, as well as misinformation about phenotypes can be entered. The openSNP project has currently no means to verify the validity of data uploaded by users. Of course, users can always delete their data or contact the team to delete stored data. Old backups of the database are deleted so that at any given time, there are only two backups. This means that deleted data disappears from the webpage immediately and will disappear after two months in the backend where it isn't accessible to the public.

@@

\subsection*{Technical implementation of the platform}
The main platform is implemented using the web framework Ruby on Rails 3.2.13. Postgres 9.2 is used as the main database backend for Rails. 
The database stores genotyping results, users' phenotypic information, literature results from Mendeley and the Public Library of Science as well as summaries on SNPs 
which can be found in SNPedia. The literature database of Mendeley is queried using the REST API, which delivers results in JSON. The literature database of 
the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format. Summaries on SNPs are provided by SNPedia, 
through querying the content via the MediaWiki API. The \emph{NHGRI GWAS Catalog} and the \emph{GET Evidence System} provide complete dumps in plain text formats. Those are regularly downloaded and parsed. SNPs that are described as 'Insufficiently evaluated' in the \emph{GET Evidence System} are not stored. All databases are queried or parsed using the unique identifier of each SNP as the search term.

@@

A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in Figure \ref{Figure4_label}. The source code of openSNP is 
published under the MIT license and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
We thank Dr. Manuel Corpas and Prof. David Edwards for constructive advice in grammar, spelling and structure of this study. Further thanks go to Samantha Clark, Fabian Zimmer and Dan Bolser for providing valuable feedback and feature ideas during the development and Thomas Down, author of \emph{BioDalliance}, and Rafael Jimenez, author of \emph{MyKaryoView}, for their support on implementing the Distributed Annotation System and the genome browser. For help with their APIs we are grateful to Mike Cariaso of \emph{SNPedia} and the PLOS \& Mendeley API teams. Thanks go to Misha Angrist and the anonymous reviewers for their helpful comments on this manuscript. We would especially like to thank the users of openSNP.org for their participation, their constructive criticism and bug-finding abilities and especially for sharing their genotyping and phenotype data.

@@

\begin{figure}[!ht]
	\begin{center}
		\includegraphics[scale=0.6]{25_10_2012_Graphs/new_pie.png}
	\end{center}
	\caption{
	{\bf Ratio of open access Publications.} Green pieces are open access. The \emph{NHGRI GWAS Catalog} doesn't give information about the open access status.}
	\label{oa_label}
\end{figure}

@@

\section*{Tables}
\begin{table}
\caption{Comparison of crowdsourced genetics platforms. N/A = Not Applicable, \checkmark = Present, \xmark = Absent}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
Name & Provides  & Provides & Open  & API & IRB approval & License \\ 
& annotation & data & enrollment & & & \\
\hline
SNPedia & \checkmark & \xmark & N/A & \checkmark & N/A & CC-NC-SA 3.0 \\ 
\hline
PGP & \checkmark & \checkmark & \xmark & \xmark & \checkmark & CC-BY 1.0, CC0 \\
\hline
PatientsLikeMe & \xmark & \checkmark & \checkmark & \xmark & \checkmark & Closed, CC-BY-SA 3.0 \\
\hline
openSNP & \checkmark & \checkmark & \checkmark & \checkmark & \xmark & CC-BY 3.0 \\
\hline
\end{tabular}
\label{tab:platforms}
\end{table}
%\begin{table}
%\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped. }  
%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%\hline
%& Mean \emph{Genotyped} & Mean \emph{Not Genotyped} & Mean Difference & Standard Error\\ 
%\hline
%\textbf{Motivation for sharing data in participants who are already genotyped} & & & & \\
%\hline 
%... curious & 3.82 & 2.66 & 1.159 & 0.193 \\ \hline % checked
%... want to help scientists & 4.64 & 4.18  & 0.465 & 0.128 \\ \hline % checked
%... for personal benefits & 3.77 & 3.32  & 0.448 & 0.183 \\ \hline % checked
%\textbf{Motivation for not sharing in participants who are not planning to get genotyped} & & & & \\ \hline
%... fear of discrimination & 3.09 & 4.15  & 1.06 & 0.195 \\ \hline % checked
%... breach of privacy & 3.01 & 3.68  & 0.666 & 0.211 \\ \hline % checked
%... fear of personalized advertising & 3.03 & 3.88  & 0.848 & 0.208 \\ \hline % checked
%... negative consequences for family members & 2.93 & 3.57  & 0.639 & 0.197 \\ \hline 
%\end{tabular}
%\label{tab:motivations1}
%\end{table}
%
%\begin{table}
%\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with participants who would not share their data.}
%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%\hline
%& Mean {Sharing} & Mean {Not Sharing} & Mean Difference & Standard Error \\ 
%\hline
%\textbf{Motivation for sharing genotypings in participants who would share} & & & & \\
%\hline
%... curiosity & 3.86 & 1.87 & 1.99 & 0.321 \\ \hline % checked
%... want to help science & 4.70 & 3.13 & 1.57 & 0.199 \\ \hline % checked
%... for personal benefits & 3.68 & 2.73 & 0.951 & 0.308 \\ \hline % checked
%\textbf{Motivation for sharing genotypings in participants who would not share} & & & & \\
%\hline
%... fear of discrimination & 3.21 & 4.73 & 1.52 & 0.322 \\ \hline % checked
%... fear of consequences for family members & 2.79 & 3.93 & 1.146 & 0.32 \\ \hline % checked
%... fear of personalized advertising & 3.17 & 4.29 & 1.112 & 0.357 \\ \hline % checked
%\end{tabular}
%\label{tab:motivations2}
%\end{table}
%\begin{table}[!ht]
%\caption{
%\bf{Table title}}
%\begin{tabular}{|c|c|c|}
%table information
%\end{tabular}
%\begin{flushleft}Table caption
%\end{flushleft}
%\label{tab:label}
% \end{table}
